White Wine Dataset - Explore and Summarize Data by James Tooles
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using
# in your analysis in this code chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk.
# This prevents the code from displaying in the knitted HTML output.
# You should set echo=FALSE for all code chunks in your file.

#install.packages('ggplot2', dependencies = T)
#install.packages('knitr', dependencies = T)
#install.packages('dplyr', dependencies = T)

library(ggplot2)
library(knitr)
library(dplyr)
```


```{r echo=FALSE, Load_the_Data}
# Load the Data
wine <- read.csv(file = 'wineQualityWhites.csv', head = TRUE, sep = ',')
```


# Univariate Plots Section
```{r echo=FALSE, Univariate_Plots0}
# Print info for wine df
str(wine)

# Print out head for wine df
head(wine)

# Print out summary for wine df
summary(wine)
```


The White Wine dataset consists of 13 variables, with almost 5,000 observations.


### What chemicial properties influence the quality of white wines?

### What chemicial properties influence the alcholol level of white wines?

# Univariate Analysis

### What is the structure of your dataset?
There are 4898 wines of 13 features (x, fixed.acidity, volatile.acidity, citric.acid, residual.sugar, cholorides, free.sulfur.dioxide, total.sulfur.dioxide, density, pH, sulphates, alcohol, quality). All of these variables are given as quanitavive variables.

Other observations:
* The 'quality' distrubution is approximately normal
* The median 'quality' value is 6.0

### What is/are the main feature(s) of interest in your dataset?

### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?

### Did you create any new variables from existing variables in the dataset?

### Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?



# Bivariate Plots Section
```{r echo=FALSE, Bivariate_Plots}

```

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in the dataset?

### Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?

### What was the strongest relationship you found?




# Multivariate Plots Section

```{r echo=FALSE, Multivariate_Plots}

```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. Were there features that strengthened each other in terms of looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the strengths and limitations of your model.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection
