White Wine Dataset - Explore and Summarize Data by James Tooles
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using
# in your analysis in this code chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk.
# This prevents the code from displaying in the knitted HTML output.
# You should set echo=FALSE for all code chunks in your file.

#install.packages('ggplot2', dependencies = T)
#install.packages('knitr', dependencies = T)
#install.packages('dplyr', dependencies = T)

library(ggplot2)
library(knitr)
library(dplyr)
```


```{r echo=FALSE, Load_the_Data}
# Load the Data
wine <- read.csv(file = 'wineQualityWhites.csv', head = TRUE, sep = ',')
```


# Univariate Plots Section


```{r echo=FALSE, Univariate_Plots0}
# Print info for wine df
str(wine)
 
# Print out head for wine df
head(wine)
  
# Print out summary for wine df
summary(wine)
```

The White Wine dataset consists of 13 variables, with almost 5,000 observations.


```{r echo=FALSE, Univariate_Plots1}
# Plot X histogram
ggplot(aes(x = X), data = wine) +
  geom_histogram(bins = length(wine$X)) 

# Plot X line graph with augmented binsize
ggplot(aes(x = X, y = X), data = wine) +
  geom_line()

# Get X statistical values
summary(wine$X)
```


Within the 'wine' df, the 'X' feature acts as an unique ID for each wine given within the df.


```{r echo=FALSE, Univariate_Plots2}
# Plot fixed.acidity histogram
ggplot(aes(x = fixed.acidity), data = wine) +
  geom_histogram(bins = 30)

# Plot fixed.acidity histogram with augmented histogram
ggplot(aes(x = fixed.acidity), data = wine) +
  geom_histogram(binwidth = 0.1) +
  scale_x_continuous(limit = c(4, 11), 
                     breaks = c(4, 5, 6, 7, 8, 9, 10, 11))

# Get X statistical values
summary(wine$fixed.acidity)
```


From the dataset documentation:

* fixed acidity: most acids involved with wine or fixed or nonvolatile (do not evaporate readily)

The 'fixed.acidity' feature seems to have a wide range of values, normally distrubuted around 6.8. There are some outliers within the wine dataset, which have a 'fixed.acidity' value greater than 11. The 'fixed.acidity' ranges in values from 3.8 to 14.2.


```{r echo=FALSE, Univariate_Plots3}
# Plot volatile.acidity histogram
ggplot(aes(x = volatile.acidity), data = wine) +
  geom_histogram(bins = 30)

# Plot volatile.acidity histogram with augmented histogram
ggplot(aes(x = volatile.acidity), data = wine) +
  geom_histogram(binwidth = 0.01) +
  scale_x_continuous(limit = c(0.1, 0.7), 
                     breaks = c(0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7))

# Get X statistical values
summary(wine$volatile.acidity)
```


From the dataset documentation:

* volatile acidity: the amount of acetic acid in wine, which at too high of levels can lead to an unpleasant, vinegar taste

The 'volatile.acidity' feature seems to have a long tail, being positively skewed. The second histogram cuts out the outliers on the few high end values for a more detailed inspection of the bulk of the data. The values for 'volatile.acidity' are centered around the median of 0.26, with a mean of 0.28. The values of 'volatile.acidity' range from 0.08 to 1.1.






The White Wine dataset consists of 13 variables, with almost 5,000 observations.


### What chemicial properties influence the quality of white wines?

### What chemicial properties influence the alcholol level of white wines?

# Univariate Analysis

### What is the structure of your dataset?
There are 4898 wines of 13 features (x, fixed.acidity, volatile.acidity, citric.acid, residual.sugar, cholorides, free.sulfur.dioxide, total.sulfur.dioxide, density, pH, sulphates, alcohol, quality). All of these variables are given as quanitavive variables.

Other observations:
* The 'quality' distrubution is approximately normal
* The median 'quality' value is 6.0

### What is/are the main feature(s) of interest in your dataset?

### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?

### Did you create any new variables from existing variables in the dataset?

### Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?



# Bivariate Plots Section
```{r echo=FALSE, Bivariate_Plots}

```

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in the dataset?

### Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?

### What was the strongest relationship you found?




# Multivariate Plots Section

```{r echo=FALSE, Multivariate_Plots}

```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. Were there features that strengthened each other in terms of looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the strengths and limitations of your model.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection
