White Wine Dataset - Explore and Summarize Data by James Tooles
========================================================

```{r echo=FALSE, warning = FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using
# in your analysis in this code chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk.
# This prevents the code from displaying in the knitted HTML output.
# You should set echo=FALSE for all code chunks in your file.

#install.packages('ggplot2', dependencies = T)
#install.packages('knitr', dependencies = T)
#install.packages('dplyr', dependencies = T)
#install.packages('GGally', dependencies = T)
#install.packages('RColorBrewer', dependencies = TRUE) 
#install.packages('memisc', dependencies = TRUE)

library(ggplot2)
library(knitr)
library(dplyr)
library(GGally)
library(RColorBrewer) 
library(memisc)
```


```{r echo=FALSE, warning = FALSE, Load_the_Data}
# Load the Data
wine <- read.csv(file = 'wineQualityWhites.csv', head = TRUE, sep = ',')
```


# Univariate Plots Section


```{r echo=FALSE, warning = FALSE, Univariate_Plots0}
# Print info for wine df
str(wine)
 
# Print out head for wine df
head(wine)
  
# Print out summary for wine df
summary(wine)
```

The White Wine dataset consists of 13 variables, with almost 5,000 observations.


```{r echo=FALSE, warning = FALSE, Univariate_Plots1}
# Drop 'X' feature from dataset
wine <- subset(wine, select = -c(X))

# Created categorical variable for 'quality'
wine$quality.cat <- as.factor(wine$quality)
```


Within the 'wine' df, the 'X' feature acts as a unique ID for each wine given within the df. X was removed from the dataframe.

Created categorical 'quality' feature named 'quality.cat'. This feature will be used in future analysis of the data.


```{r echo=FALSE, warning = FALSE, Univariate_Plots2}
# Plot fixed.acidity histogram
ggplot(aes(x = fixed.acidity), data = wine) +
  geom_histogram(bins = 30)

# Plot fixed.acidity histogram with augmented histogram
ggplot(aes(x = fixed.acidity), data = wine) +
  geom_histogram(binwidth = 0.1) +
  scale_x_continuous(limit = c(4, 11), 
                     breaks = c(seq(4, 11, by = 0.5)))

# Get fixed.acidity statistical values
summary(wine$fixed.acidity)
```


From the dataset documentation:

* fixed acidity (tartaric acid - g / dm^3): most acids involved with wine or fixed or nonvolatile (do not evaporate readily)

The 'fixed.acidity' feature seems to have a wide range of values, normally distributed around 6.80. There are some outliers within the wine dataset, which have a 'fixed.acidity' value greater than 11. The 'fixed.acidity' ranges in values from 3.80 to 14.20.


```{r echo=FALSE, warning = FALSE, Univariate_Plots3}
# Plot volatile.acidity histogram
ggplot(aes(x = volatile.acidity), data = wine) +
  geom_histogram(bins = 30)

# Plot volatile.acidity histogram with augmented histogram
ggplot(aes(x = volatile.acidity), data = wine) +
  geom_histogram(binwidth = 0.01) +
  scale_x_continuous(limit = c(0.1, 0.7), 
                     breaks = c(seq(0.1, 0.7, by = 0.1)))

# Get volatile.acidity statistical values
summary(wine$volatile.acidity)
```


From the dataset documentation:

* volatile acidity (acetic acid - g / dm^3): the amount of acetic acid in wine, which at too high of levels can lead to an unpleasant, vinegar taste

The 'volatile.acidity' feature seems to have a long tail, being positively skewed. The second histogram cuts out the outliers on the few high end values for a more detailed inspection of the bulk of the data. The values for 'volatile.acidity' are centered around the median of 0.26, with a mean of 0.28. The values of 'volatile.acidity' range from 0.08 to 1.10. Based on the description given by the dataset documentation, keeping the 'volatile.acidity' values low would be in the best interest of the consumer.


```{r echo=FALSE, warning = FALSE, Univariate_Plots4}
# Plot citric.acid histogram
ggplot(aes(x = citric.acid), data = wine) +
  geom_histogram(bins = 30)

# Plot citric.acid histogram with augmented histogram
ggplot(aes(x = citric.acid), data = wine) +
  geom_histogram(binwidth = 0.01) +
  scale_x_continuous(limit = c(0.1, 0.8), 
                     breaks = c(seq(0.1, 0.8, by = 0.1)))

# Get citric.acid statistical values
summary(wine$citric.acid)
```


From the dataset documentation:

* citric acid (g / dm^3): found in small quantities, citric acid can add 'freshness' and flavor to wines

The 'citric.acid' feature seems to have a long tail, bring positively skewed. The values for 'citric.acid' are centered around the median of 0.32, with a mean of 0.33. The values of 'citric.acid' range from 0.00 to 1.66. There is a huge spike in the count of 'citric.acid' at 0.5, this could possibly an industry agreed "sweet spot". Based on the description given by the dataset documentation, keeping the 'citric acid' values high would be in the best interest of the consumer.


```{r echo=FALSE, warning = FALSE, Univariate_Plots5}
# Plot residual.sugar histogram
ggplot(aes(x = residual.sugar), data = wine) +
  geom_histogram(bins = 30)

# Plot residual.sugar histogram with limits
ggplot(aes(x = residual.sugar), data = wine) +
  geom_histogram(binwidth = 0.1) +
  scale_x_continuous(limits = c(0, 21),
                     breaks = c(seq(0, 21, by = 1)))

# Get residual.sugar statistical values
summary(wine$residual.sugar)
```

From the dataset documentation:

* residual sugar (g / dm^3): the amount of sugar remaining after fermentation stops, it's rare to find wines with less than 1 gram/liter and wines with greater than 45 grams/liter are considered sweet

The 'residual.sugar' feature seems to have a long tail, being positively skewed. The values for residual sugar have high concentrations and spike around the values of 1 and 2, later dropping off. The median for residual sugar is 5.20, with a mean of 6.39.


```{r echo=FALSE, warning = FALSE, Univariate_Plots6}
# Plot chlorides histogram
ggplot(aes(x = chlorides), data = wine) +
  geom_histogram(bins = 30)

# Plot chlorides histogram with limits
ggplot(aes(x = chlorides), data = wine) +
  geom_histogram(binwidth = 0.001) +
  scale_x_continuous(limits = c(0, 0.2),
                     breaks = c(seq(0, 0.2, by = 0.01)))

# Get residual.sugar statistical values
summary(wine$chlorides)
```


From the dataset documentation:

* chlorides (sodium chloride - g / dm^3): the amount of salt in the wine

The 'chlorides' feature looks normally distributed around the median of 0.043, with a mean of 0.046 but, the data does have a long tail to the right that makes up a small portion of the overall wines. It would make sense for wines to have the smallest amount of salt possible due to wines not being known for their high salinity.


```{r echo=FALSE, warning = FALSE, Univariate_Plots7}
# Plot free.sulfur.dioxide histogram
ggplot(aes(x = free.sulfur.dioxide), data = wine) +
  geom_histogram(bins = 30)

# Plot free.sulfur.dioxide histogram with limits
ggplot(aes(x = free.sulfur.dioxide), data = wine) +
  geom_histogram(binwidth = 1) +
  scale_x_continuous(limits = c(0, 100),
                     breaks = c(seq(0, 100, by = 5)))

# Get free.sulfur.dioxide statistical values
summary(wine$free.sulfur.dioxide)
```


From the dataset documentation:

* free sulfur dioxide (mg / dm^3): the free form of SO2 exists in equilibrium between molecular SO2 (as a dissolved gas) and bisulfite ion; it prevents microbial growth and the oxidation of wine

The 'free.sulfur.dioxide' feature looks normally distributed around the median 34.00, with a mean of 35.31. The histogram has a long tail to the right, showing outliers with extreme free sulfur dioxide levels. Based on the description of from the dataset documentation, this feature acts as a preservative for the wine.


```{r echo=FALSE, warning = FALSE, Univariate_Plots8}
# Plot total.sulfur.dioxide histogram
ggplot(aes(x = total.sulfur.dioxide), data = wine) +
  geom_histogram(bins = 30)

# Plot total.sulfur.dioxide histogram with limits
ggplot(aes(x = total.sulfur.dioxide), data = wine) +
  geom_histogram(binwidth = 1) +
  scale_x_continuous(limits = c(0, 300),
                     breaks = c(seq(0, 300, by = 20)))

# Get total.sulfur.dioxide statistical values
summary(wine$total.sulfur.dioxide)
```


From the dataset documentation:

* total sulfur dioxide (mg / dm^3): amount of free and bound forms of S02; in low concentrations, SO2 is mostly undetectable in wine, but at free SO2 concentrations over 50 ppm, SO2 becomes evident in the nose and taste of wine

The 'total.sulfur.dioxide' feature looks normally distributed around the median 134.00, with a mean of 138.40. The histogram has a small tail to the right, showing outliers with extreme total sulfur dioxide levels. Based on the description of from the dataset documentation, the taste of the wine can be affected if the concentration of total sulfur dioxide becomes too high.


```{r echo=FALSE, warning = FALSE, Univariate_Plots9}
# Plot density histogram
ggplot(aes(x = density), data = wine) +
  geom_histogram(bins = 30)

# Plot density histogram with limits
ggplot(aes(x = density), data = wine) +
  geom_histogram(binwidth = 0.0001) +
  scale_x_continuous(limits = c(0.985, 1.005),
                     breaks = c(seq(0.985, 1.005, by = 0.002)))

# Get density statistical values
summary(wine$density)
```


From the dataset documentation:

* density (g / cm^3): the density of water is close to that of water depending on the percent alcohol and sugar content

The 'density' feature looks normally distributed around the median 0.9937, with a mean of 0.9940. The values for wine density within the dataset has some high value outliers, that were cut off when inspecting the second 'density' histogram. Based on the description from the dataset documentation, the wine density is a variable dependent upon the values of alcohol and sugar content.


```{r echo=FALSE, warning = FALSE, Univariate_Plots10}
# Plot pH histogram
ggplot(aes(x = pH), data = wine) +
  geom_histogram(bins = 30)

# Plot pH histogram with smaller binwidth
ggplot(aes(x = pH), data = wine) +
  geom_histogram(binwidth = 0.01) +
  scale_x_continuous(breaks = c(seq(2.7, 3.8, by = 0.1)))

# Get pH statistical values
summary(wine$pH)
```

From the dataset documentation:

* pH: describes how acidic or basic a wine is on a scale from 0 (very acidic) to 14 (very basic); most wines are between 3-4 on the pH scale

The 'pH' level is known to correlate with the taste, from bitter (high pH = basic) to sour (low pH = acidic). The first 'pH' histogram shows a nice Gaussian distribution of the values for 'pH', centered around the median 3.18, with a mean of 3.188. This is the same shape as the 'quality' feature histogram. The more defined, second histogram, shows some interesting spikes in the counts for different 'pH' levels throughout the histogram.


```{r echo=FALSE, warning = FALSE, Univariate_Plots11}
# Plot sulphates histogram
ggplot(aes(x = sulphates), data = wine) +
  geom_histogram(bins = 30)

# Plot sulphates histogram with smaller binwidth
ggplot(aes(x = sulphates), data = wine) +
  geom_histogram(binwidth = 0.01) +
  scale_x_continuous(limits = c(0.2, 1.0),
                     breaks = c(seq(0.2, 1.0, by = 0.05)))

# Get sulphates statistical values
summary(wine$sulphates)
```


From the dataset documentation:

* sulphates (potassium sulphate - g / dm3): a wine additive which can contribute to sulfur dioxide gas (S02) levels, which acts as an antimicrobial and antioxidant

The 'sulphates' feature looks normally distributed around the median 0.4700, with a mean of 0.4898. The histogram has a long tail to the right, showing outliers with extreme free sulfur dioxide levels. Based on the description of from the dataset documentation, this feature acts as a preservative for the wine.


```{r echo=FALSE, warning = FALSE, Univariate_Plots12}
# Plot alcohol histogram
ggplot(aes(x = alcohol), data = wine) +
  geom_histogram(bins = 30)

# Plot alcohol histogram with cut off lower outliers
ggplot(aes(x = alcohol), data = wine) +
  geom_histogram(binwidth = 0.1) +
  scale_x_continuous(limit = c(8.0, 14.2),
                     breaks = c(seq(8.0, 14.2, by = 0.5)))

# Get alcohol statistical values
summary(wine$alcohol)
```


From the dataset documentation:

* alcohol (% by volume): the percent alcohol content of the wine

At the end of the day, wine is an alcoholic product. Perhaps the 'alcohol' content has something to do with the 'quality' value of a wine. The first histogram shows the raw histogram of the 'alcohol' feature. There seem to be a small number of lower outliers.

The second histogram of the 'alcohol' feature removes theses smaller 'alcohol' values from the histogram. Both histograms are positively skewed. There is a noticeable spike at wine alcohol value 9.5. The median of wine alcohol is 10.40, with a mean of 10.51.


```{r echo=FALSE, warning = FALSE, Univariate_Plots13}
# Plot quality bar graph
ggplot(aes(x = quality.cat), data = wine) +
  geom_bar()

# Plot quality bar graph with augmentations
ggplot(aes(x = quality), data = wine) +
  geom_bar() +
  scale_x_continuous(limit = c(0, 10),
                     breaks = c(seq(0, 10, by = 1)))

# Get quality statistical values
summary(wine$quality)
```


From the dataset documentation:

* quality (score between 0 and 10)

Starting with a default histogram of the 'quality' feature within the 'wine' dataset resembles a Gaussian distribution of the 'quality' feature, centered at a 'quality' value of 6. By increasing the bin size of the histogram allowed a better visual depiction of the dataset, which is slightly positively skewed. These graphs support the findings within the statistical summary for 'quality' with a median value of 6, and a mean value of 5.878.

The dataset documentation states that the 'quality' feature should range from 0-10, unfortunately all of the wines within this dataset range from 3-9.


# Univariate Analysis


### What is the structure of your dataset?
There are 4898 wines of 13 features (x, fixed.acidity, volatile.acidity, citric.acid, residual.sugar, chlorides, free.sulfur.dioxide, total.sulfur.dioxide, density, pH, sulphates, alcohol, quality). All of these variables are given as quantitative variables.

* **X**: a unique id identifier for each wine

* **fixed acidity**: most acids involved with wine or fixed or nonvolatile (do not evaporate readily)

* **volatile acidity**: the amount of acetic acid in wine, which at too high of levels can lead to an unpleasant, vinegar taste

* **citric acid**: found in small quantities, citric acid can add 'freshness' and flavor to wines

* **residual sugar**: the amount of sugar remaining after fermentation stops, it's rare to find wines with less than 1 gram/liter and wines with greater than 45 grams/liter are considered sweet

* **chlorides**: the amount of salt in the wine

* **free sulfur dioxide**: the free form of SO2 exists in equilibrium between molecular SO2 (as a dissolved gas) and bisulfite ion; it prevents microbial growth and the oxidation of wine

* **total sulfur dioxide**: amount of free and bound forms of S02; in low concentrations, SO2 is mostly undetectable in wine, but at free SO2 concentrations over 50 ppm, SO2 becomes evident in the nose and taste of wine

* **density**: the density of water is close to that of water depending on the percent alcohol and sugar content

* **pH**: describes how acidic or basic a wine is on a scale from 0 (very acidic) to 14 (very basic); most wines are between 3-4 on the pH scale

* **sulphates**: a wine additive which can contribute to sulfur dioxide gas (S02) levels, which acts as an antimicrobial and antioxidant

* **alcohol**: the percent alcohol content of the wine

* **quality** (score between 0 and 10)
    * (worst) 0 -------> 10 (best)

Other observations:

* The 'quality' distribution is approximately normal
* The median 'quality' value is 6.0
* Citric acid has a spike in wine count at a value of 0.5

### What are the main features of interest in your dataset?
The main features of interest are any features that significantly contribute to the quality output value. The goal of this analysis is to determine which features are best for predicting the quality of a wine.

### What other features in the dataset do you think will help support your investigation into your feature of interest?
Volatile acidity, citric acid, residual sugar, chlorides, total sulfur dioxide, pH, and alcohol. All of these features are described in the dataset documentation as to affect the taste, smell, or other measurements of value to consumers such as alcohol content.

The other features such as, fixed acidity, free sulfur dioxide, density, sulphates, do not seem to play a role in determining quality value due to the description given within the dataset documentation. All of these features are either dependent upon other features of interest, preservatives, or do not show enough variance to be heavily weighted in determining wine quality.

### Did you create any new variables from existing variables in the dataset?
Yes, a categorical feature of quality was created called 'quality.cat'.

### Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?
The citric acid feature has a spike in wine count at a value of 0.5, this seems odd and interesting and will be investigated for any impact in wine quality within the bivariate section.

All of the features were limited, outliers removed, and/or parsed in more visual detail to see any hidden attributes within the feature.

The quality feature, was shown on a bar graph to highlight the ordinal property of the feature.



# Bivariate Plots Section


```{r echo=FALSE, warning = FALSE, Bivariate_Plots0}
# Get correlation matrix values
wine_numeric <- subset(wine, select = -c(quality.cat))
round(cor(wine_numeric), 2)

#Set the seed for reproducible results
set.seed(0)

# Get correlation matrix graph
theme_set(theme_bw(5))
names(wine_numeric)
ggpairs(wine_numeric[sample.int(nrow(wine_numeric), 500), ])
```


From a subset of the citric acid, residual sugar, and pH do not seem to have strong correlations with quality, but volatile acidity, chlorides, total sulfur dioxide, density, and alcohol are moderately correlated with quality. I want to look closer at scatter plots involving quality and some other variables like volatile acidity, chlorides, density, and alcohol.


```{r echo=FALSE, warning = FALSE, Bivariate_Plots1}
#Reset ggplot theme and text size
theme_set(theme_minimal(9))

# Plot quality scatter quality vs volatile.acidity
ggplot(aes(x = volatile.acidity, y = quality), data = wine) +
  geom_point(alpha = 0.5, position = position_jitter(width = 0, height = 0.5)) +
  geom_smooth(method = 'lm', color = 'red') +
  scale_x_continuous(breaks = c(seq(0, 1.10, by = 0.1))) +
  scale_y_continuous(breaks = c(seq(0, 10, by = 1)), 
                     limit = c(3, 9)) 

# Plot quality boxplot volatile.acidity vs quality
ggplot(aes(x = quality.cat, y = volatile.acidity, colour = quality.cat), data = wine) +
  geom_boxplot() +
  scale_y_continuous(limit = c(0, 0.8),
                     breaks = c(seq(0, 0.8, by = 0.1)))

# Get correlation value
cor.test(wine$quality, wine$volatile.acidity)
```


The high quality wines have a low volatile acidity value, only the sub-par wines have high volatile acidity values. This analysis only tells us part of the story given the correlation value of -0.195 for quality vs volatile acidity.



```{r echo=FALSE, warning = FALSE, Bivariate_Plots2}
# Plot quality scatter quality vs chlorides
ggplot(aes(x = chlorides, y = quality), data = wine) +
  geom_point(alpha = 0.5, position = position_jitter(width = 0, height = 0.5)) +
  geom_smooth(method = 'lm', color = 'red') +
  scale_x_continuous(limits = c(0, 0.2),
                     breaks = c(seq(0, 0.2, by = 0.01))) +
  scale_y_continuous(breaks = c(seq(0, 10, by = 1)), 
                     limit = c(3, 9)) 

# Plot quality boxplot chlorides vs quality
ggplot(aes(x = quality.cat, y = chlorides, colour = quality.cat), data = wine) +
  geom_boxplot() +
  scale_y_continuous(limits = c(0, 0.15),
                     breaks = c(seq(0, 0.15, by = 0.01)))

# Get correlation value
cor.test(wine$quality, wine$chlorides)
```


The majority of wines have low chlorides values. The high quality wines have a low chlorides value; a large number of average wines have high chlorides values. There are a few outliers on the high quality and low quality wines with a large chlorides value. This analysis only tells us part of the story given the correlation value of -0.210 for quality vs chlorides.


```{r echo=FALSE, warning = FALSE, Bivariate_Plots3}
# Plot quality scatter quality vs total.sulfur.dioxide
ggplot(aes(x = total.sulfur.dioxide, y = quality), data = wine) +
  geom_point(alpha = 0.5, position = position_jitter(width = 0, height = 0.5)) +
  geom_smooth(method = 'lm', color = 'red') +
  scale_x_continuous(limits = c(0, 300),
                     breaks = c(seq(0, 300, by = 20))) +
  scale_y_continuous(breaks = c(seq(0, 10, by = 1)), 
                     limit = c(3, 9)) 

# Plot quality boxplot total.sulfur.dioxide vs quality
ggplot(aes(x = quality.cat, y = total.sulfur.dioxide, colour = quality.cat), data = wine) +
  geom_boxplot() +
  scale_y_continuous(limits = c(0, 300),
                     breaks = c(seq(0, 300, by = 20)))

# Get correlation value
cor.test(wine$quality, wine$total.sulfur.dioxide)
```


As the quality of the wines increase, the limits for total sulfur dioxide values seem to converge from 80 to 140, with the low quality wines having a much larger variation of total sulfur dioxide values, about 20 to 280. This analysis only tells us part of the story given the correlation value of -0.175 for quality vs total sulfur dioxide.


```{r echo=FALSE, warning = FALSE, Bivariate_Plots4}
# Plot quality scatter quality vs density
ggplot(aes(x = density, y = quality), data = wine) +
  geom_point(alpha = 0.5, position = position_jitter(width = 0, height = 0.5)) +
  geom_smooth(method = 'lm', color = 'red') +
  scale_x_continuous(limits = c(0.985, 1.005),
                     breaks = c(seq(0.985, 1.005, by = 0.002))) +
  scale_y_continuous(breaks = c(seq(0, 10, by = 1)), 
                     limit = c(3, 9)) 

# Plot quality boxplot density vs quality
ggplot(aes(x = quality.cat, y = density, colour = quality.cat), data = wine) +
  geom_boxplot() +
  scale_y_continuous(limits = c(0.985, 1.005),
                     breaks = c(seq(0.985, 1.005, by = 0.002)))

# Get correlation value
cor.test(wine$quality, wine$density)
```


The high quality wines have a low density value, only the sub-par wines have high density values. The density and quality of wine has a decent correlation value of -0.307. 



```{r echo=FALSE, warning = FALSE, Bivariate_Plots5}
# Plot quality scatter quality vs alcohol
ggplot(aes(x = alcohol, y = quality), data = wine) +
  geom_point(alpha = 0.5, position = position_jitter(width = 0.1, height = 0.5)) +
  geom_smooth(method = 'lm', color = 'red') +
  scale_x_continuous(limit = c(8.0, 14.2),
                     breaks = c(seq(8.0, 14.2, by = 0.5))) +
  scale_y_continuous(breaks = c(seq(0, 10, by = 1)), 
                     limit = c(3, 9)) 

# Plot quality boxplot alcohol vs quality
ggplot(aes(x = quality.cat, y = alcohol, colour = quality.cat), data = wine) +
  geom_boxplot() +
  scale_y_continuous(limit = c(8.0, 14.2),
                     breaks = c(seq(8.0, 14.2, by = 0.5)))

# Get correlation value
cor.test(wine$quality, wine$alcohol)
```


The high quality wines have a higher alcohol value, only the sub-par wines have lower alcohol values. Only a small portion of wines have high quality and low alcohol content. The quality and alcohol features have the highest correlation yet, at a value of 0.436.


```{r echo=FALSE, warning = FALSE, Bivariate_Plots6}
# Plot scatter density vs alcohol
ggplot(aes(x = alcohol, y = density), data = wine) +
  geom_point(alpha = 0.5, position = position_jitter(width = 0.1)) +
  geom_smooth(method = 'lm', color = 'red') +
  scale_x_continuous(limit = c(8.0, 14.2),
                     breaks = c(seq(8.0, 14.2, by = 0.5))) +
  scale_y_continuous(limits = c(0.985, 1.005),
                     breaks = c(seq(0.985, 1.005, by = 0.002)))  

# Get correlation value
cor.test(wine$density, wine$alcohol)
```

As the alcohol values increase, the density decreases. This could be due to alcohol having a lower density than the rest of the wine content, therefore bringing down the overall density of the wine. Alcohol and density have a strong negative correlation value of -0.780.

```{r echo=FALSE, warning = FALSE, Bivariate_Plots7}
# Plot scatter density vs residual.sugar
ggplot(aes(x = residual.sugar, y = density), data = wine) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = 'lm', color = 'red') +
  scale_x_continuous(limits = c(0, 21),
                     breaks = c(seq(0, 21, by = 1))) +
  scale_y_continuous(limits = c(0.985, 1.005),
                     breaks = c(seq(0.985, 1.005, by = 0.002)))  

# Get correlation value
cor.test(wine$density, wine$residual.sugar)
```


The residual sugar and density have a strong positive correlation of 0.839. As the residual sugar increases, the density increases. This could be due to the fact that residual sugar has a higher density compared to the remaining wine content. This could also be related to the fact that during fermentation, sugar is turned into alcohol. So the less sugar turned into lower density alcohol, the higher the density of the wine.


```{r echo=FALSE, warning = FALSE, Bivariate_Plots8}
# Plot scatter alcohol vs residual.sugar
ggplot(aes(x = residual.sugar, y = alcohol), data = wine) +
  geom_point(alpha = 0.5, position = position_jitter(width = 0, height = 0.1)) +
  geom_smooth(method = 'lm', color = 'red') +
  scale_x_continuous(limits = c(0, 21),
                     breaks = c(seq(0, 21, by = 1))) +
  scale_y_continuous(limit = c(8.0, 14.2),
                     breaks = c(seq(8.0, 14.2, by = 0.5))) 

# Get correlation value
cor.test(wine$alcohol, wine$residual.sugar)
```

The alcohol content and residual.sugar have a moderate negative correlation of -0.451. As the residual sugar increases, the alcohol content decreases. This is most likely due to the fermination process of making acohol, which consumes sugar to produce alcohol.


# Bivariate Analysis


### Talk about some of the relationships you observed in this part of the investigation. How did the features of interest vary with other features in the dataset?
The relationship between volatile acidity and quality of wines has a parabolic shape, in the sense that the low and high quality wines have a higher level of volatile acidity, while the average wines have a lower volatile acidity.

For the chlorides levels, the chlorides values decrease as the quality of wine increases.

The total sulfur dioxide has one of the most interesting relationships to quality, as it converges to a total sulfur dioxide range of 80 to 140 as the quality of wine increases. The lower quality wines have a much larger variation of total sulfur dioxide values.

The wine density decreases as the wine quality increase, with a significant drop as the wine quality equals 9.

As the median alcohol level increases, the wine quality increases. The alcohol feature has the highest correlation with wine quality out of all of the given features.

### Did you observe any interesting relationships between the other features (not the main features) of interest)?
The density, alcohol, and residual sugar features seem to be highly correlated with each other. The alcohol is created by consuming sugar during the fermentation process, and the more alcohol within the wine the lower the density of the wine given the density of the alcohol relative to residual sugar and the remaining wine content.

### What was the strongest relationship you found?
The relationships between the wine features of density and alcohol, alcohol and residual sugar, residual sugar and alcohol, and alcohol and quality.


# Multivariate Plots Section


```{r echo=FALSE, warning = FALSE, Multivariate_Plots0}

# Plot scatter density vs alcohol with quality (colour)
theme_set(theme_dark(9))
ggplot(aes(x = alcohol, y = density, colour = quality.cat), data = wine) +
  geom_point(position = position_jitter(width = 0.2, height = 0)) +
  scale_color_brewer(type = 'div',
                     guide = guide_legend(title = 'Quality', reverse = TRUE,
                                          override.aes = list(alpha = 1, size = 2))) +
  scale_x_continuous(limit = c(8.0, 14.2),
                     breaks = c(seq(8.0, 14.2, by = 0.5))) +
  scale_y_continuous(limits = c(0.985, 1.005),
                     breaks = c(seq(0.985, 1.005, by = 0.002)))


# Plot scatter density vs alcohol with quality (colour - separated) 
ggplot(aes(x = alcohol, y = density, colour = quality.cat), data = wine) +
  geom_point(position = position_jitter(width = 0.2, height = 0)) +
  scale_color_brewer(type = 'div',
                     guide = guide_legend(title = 'Quality', reverse = TRUE,
                                          override.aes = list(alpha = 1, size = 2))) +
  scale_x_continuous(limit = c(8.0, 14.2),
                     breaks = c(seq(8.0, 14.2, by = 0.5))) +
  scale_y_continuous(limits = c(0.985, 1.005),
                     breaks = c(seq(0.985, 1.005, by = 0.002))) +
  facet_wrap(~quality.cat)
```


The first graph shows us the quality distribution as a function of color, plotted in a density vs alcohol graph. The graph is then separated to show the different trends per quality value. The wine quality starts at a value of 3, where the wines are grouped around high density and low alcohol values. As the wine quality increases, the points tend towards higher alcohol levels and lower density levels.



```{r echo=FALSE, warning = FALSE, Multivariate_Plots1}
# Plot scatter density vs residual.sugar with quality (colour)
ggplot(aes(x = residual.sugar, y = density, colour = quality.cat), data = wine) +
  geom_point(position = position_jitter(width = 0.2, height = 0)) +
  scale_color_brewer(type = 'div',
                     guide = guide_legend(title = 'Quality', reverse = TRUE,
                                          override.aes = list(alpha = 1, size = 2))) +
  scale_x_continuous(limits = c(0, 21),
                     breaks = c(seq(0, 21, by = 1))) +
  scale_y_continuous(limits = c(0.985, 1.005),
                     breaks = c(seq(0.985, 1.005, by = 0.002)))


# Plot scatter density vs residual.sugar with quality (colour - separated) 
ggplot(aes(x = residual.sugar, y = density, colour = quality.cat), data = wine) +
  geom_point(position = position_jitter(width = 0.2, height = 0)) +
  scale_color_brewer(type = 'div',
                     guide = guide_legend(title = 'Quality', reverse = TRUE,
                                          override.aes = list(alpha = 1, size = 2))) +
  scale_x_continuous(limits = c(0, 21),
                     breaks = c(seq(0, 21, by = 1))) +
  scale_y_continuous(limits = c(0.985, 1.005),
                     breaks = c(seq(0.985, 1.005, by = 0.002))) +
  facet_wrap(~quality.cat)
```


The residual sugar and density relationships have extremely high variation, except at the highest quality level. The majority of wine quality 9, have both low residual sugar and density.



```{r echo=FALSE, warning = FALSE, Multivariate_Plots2}
# Plot scatter alcohol vs residual.sugar with quality (colour)
ggplot(aes(x = residual.sugar, y = alcohol, colour = quality.cat), data = wine) +
  geom_point(position = position_jitter(width = 0.2, height = 0.1)) +
  scale_color_brewer(type = 'div',
                     guide = guide_legend(title = 'Quality', reverse = TRUE,
                                          override.aes = list(alpha = 1, size = 2))) +
  scale_x_continuous(limits = c(0, 21),
                     breaks = c(seq(0, 21, by = 1))) +
  scale_y_continuous(limit = c(8.0, 14.2),
                     breaks = c(seq(8.0, 14.2, by = 0.5)))


# Plot scatter alcohol vs residual.sugar with quality (colour - separated) 
ggplot(aes(x = residual.sugar, y = alcohol, colour = quality.cat), data = wine) +
  geom_point(position = position_jitter(width = 0.2, height = 0.1)) +
  scale_color_brewer(type = 'div',
                     guide = guide_legend(title = 'Quality', reverse = TRUE,
                                          override.aes = list(alpha = 1, size = 2))) +
  scale_x_continuous(limits = c(0, 21),
                     breaks = c(seq(0, 21, by = 1))) +
  scale_y_continuous(limit = c(8.0, 14.2),
                     breaks = c(seq(8.0, 14.2, by = 0.5))) +
  facet_wrap(~quality.cat)
```


The first graph shows us the quality distribution as a function of color, plotted in a alcohol vs residual sugar graph. The graph is then separated to show the different trends per quality value. The wine quality starts at a value of 3, where the wines are grouped around low alcohol and a variation of residual sugar values. As the wine quality increases, the points tend towards higher alcohol levels and lower residual sugar levels.


```{r echo=FALSE, warning = FALSE, Multivariate_Plots3}
#Create Linear Models
m1 <- lm(quality ~ fixed.acidity, data = wine)
m2 <- update(m1, ~ . + volatile.acidity)
m3 <- update(m2, ~ . + citric.acid)
m4 <- update(m3, ~ . + residual.sugar)
m5 <- update(m4, ~ . + chlorides)
m6 <- update(m5, ~ . + free.sulfur.dioxide)
m7 <- update(m6, ~ . + total.sulfur.dioxide)
m8 <- update(m7, ~ . + density)
m9 <- update(m8, ~ . + pH)
m10 <- update(m9, ~ . + sulphates)
m11 <- update(m10, ~ . + alcohol)
mtable(m1, m2, m3, m4, m5, m6, m7, m8, m9, m10, m11, sdigits = 3)
```


The linear model created has extremely low R-squared values. This could be due to the diversity of the dataset given in the original wine dataset. The majority of the data given has a wine quality value of 5, 6, 7. Shown in the quality bar graph in the Univariate section. Having a more diverse dataset would lead to a better modeling of the data.


# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. Were there features that strengthened each other in terms of looking at your feature of interest?
The density vs alcohol graph showing the quality of the wines was insightful. There seemed to be an overall trend, as you can see the shifting of plot points as the quality of wine increases.

The density vs residual sugar graph showing the quality of the wines, highlighted the variation of wines with low wine quality, while the high quality wines seemed to converge around a similar density/residual sugar ratio.

The alcohol vs residual sugar graph was a mix between the trends shown in the first two graphs in reference to the wine quality values. As the wine quality increased, the movement trend of the alcohol/residual sugar ratio was shown, as well as a decrease in variation of the feature values as the wine quality increased.

### Were there any interesting or surprising interactions between features?
The general decrease in variation of the observed features as the wine quality increased was interesting. From this dataset it showed that as the wine quality increases the wine features converge to a given range of feature values, for the observed features.


### OPTIONAL: Did you create any models with your dataset? Discuss the strengths and limitations of your model.

Yes, a linear model was created for this dataset. The reported R-squared values are extremely low. This is most likely due to the lack of variation within the dataset for the wine quality values. This dataset does not fair well for predictive linear models, when the output feature (wine quality) is so concentrated in <30% of the possible wine quality values.

------


# Final Plots and Summary


### Plot One
```{r echo=FALSE, warning = FALSE, Plot_One}
# Plot final quality bar graph
ggplot(aes(x = quality.cat, fill = quality.cat), data = wine) +
  geom_bar() +
  scale_y_continuous(breaks = c(seq(0, 2300, by = 250))) +
  labs(title = 'Wine Dataset Quality Feature Distribution',
       x = 'Wine Quality', 
       y = 'Frequency Count') +
  guides(fill = guide_legend(title="Quality", 
                             fill = wine$quality.cat, 
                             reverse = TRUE)) +
  theme(plot.title = element_text(size=20, face="bold", 
    margin = margin(10, 0, 10, 0)))

```


### Description One
The distribution of wine qualities appears to be normal, centered around 6 but, due to the lack of variation in the wine quality feature, it was difficult to determine what contributes to a high quality or low quality wine, due to the majority of wines in this dataset having an average quality.


### Plot Two


```{r echo=FALSE, warning = FALSE, Plot_Two}
# Plot final scatter density vs alcohol with quality (colour)
ggplot(aes(x = alcohol, y = density, colour = quality.cat), data = wine) +
  geom_point(alpha = 0.8, position = position_jitter(width = 0.2, height = 0)) +
  scale_x_continuous(limit = c(8.0, 14.2),
                     breaks = c(seq(8.0, 14.2, by = 0.5))) +
  scale_y_continuous(limits = c(0.987, 1.003),
                     breaks = c(seq(0.987, 1.003, by = 0.002))) +
  labs(title = 'Wine Dataset Quality Feature Distribution',
       x = 'Wine Alcohol Level [% by volume]', 
       y = 'Wine Density [g / cm^3]') +
  scale_color_brewer(type = 'div',
                     guide = guide_legend(title = 'Quality', reverse = TRUE,
                                          override.aes = list(alpha = 1, size = 2))) +
  theme(plot.title = element_text(size=20, face="bold", 
    margin = margin(10, 0, 10, 0)))
```


### Description Two
The average wine quality is over represented within this wine dataset but with the above graph it is clear that the majority of high quality wines tend to have high alcohol levels and low density levels.


### Plot Three


```{r echo=FALSE, warning = FALSE, Plot_Three}
# Plot final quality boxplot alcohol vs quality
theme_set(theme_minimal(9))

ggplot(aes(x = quality.cat, y = alcohol, fill = quality.cat), data = wine) +
  geom_boxplot() +
  scale_y_continuous(limit = c(8.0, 14.2),
                     breaks = c(seq(8.0, 14.2, by = 0.5))) +
  labs(title = 'Wine Dataset Alcohol Level Per Wine Quality',
       x = 'Wine Quality', 
       y = 'Wine Alcohol Level [% by volume]') +
  guides(fill = guide_legend(title="Quality", 
                             colour = wine$quality.cat, 
                             reverse = TRUE)) +
  theme(plot.title = element_text(size=20, face="bold", 
    margin = margin(10, 0, 10, 0)))
```


### Description Three
The alcohol level of a wine by volume tends to be positively correlated and increase with the wine quality level. Generally, in the dataset given, the higher the median alcohol value, the higher the quality of wine.


------


# Reflection
The white wine data set contains information on almost 5,000 wines, with 12 different features, and was created in 2009. I started by observing the distributions of each feature within the dataset. From there I took note of any trends, abnormalities, and possible insights into the meanings behind the features. I then started cross plotting and correlating the different features with quality, and some of the features with other input features given in the dataset. Finally, I explored the additive effects of features on the overall wine quality feature and created a linear model to better understand the dataset.

There was a clear trend with median alcohol content and wine quality. There were also clear trends with a combination of residual sugars, alcohol, density, and wine quality. Some of these variables could be dependent upon each other, more domain knowledge is needed for further insight into the data exploration. The relationships between the afore mentioned features became more clear, when scatter plots were created, and separated by wine quality. The plot points migrated, converged, and created a clear underlying relationship to the wine quality.

Unfortunately, the dataset given is heavily concentrated on average wine qualities. The lowest, and highest wine quality levels were sparse within this dataset. Giving way to a poor predictive linear model, with a low R-squared value, due to the lack of variation in the wine quality (output feature). This data could also be non-linear, feature incomplete, and otherwise not verbose enough to gain a more definitive model. Other machine learning models were not attempted, and a linear model might not be the best approach to modeling this dataset. For future work, a wider range of features for the wine, such as, price, age, location of production, and other features would greatly enrich the dataset. In the end, a median high alcohol percentage wine, gave way to a higher quality wine, other features could and do impact the wine quality value but, this feature trended strongly given the final box plot shown within this exploration.

