White Wine Dataset - Explore and Summarize Data by James Tooles
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using
# in your analysis in this code chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk.
# This prevents the code from displaying in the knitted HTML output.
# You should set echo=FALSE for all code chunks in your file.

#install.packages('ggplot2', dependencies = T)
#install.packages('knitr', dependencies = T)
#install.packages('dplyr', dependencies = T)

library(ggplot2)
library(knitr)
library(dplyr)
```


```{r echo=FALSE, Load_the_Data}
# Load the Data
wine <- read.csv(file = 'wineQualityWhites.csv', head = TRUE, sep = ',')
```


# Univariate Plots Section


```{r echo=FALSE, Univariate_Plots0}
# Print info for wine df
str(wine)
 
# Print out head for wine df
head(wine)
  
# Print out summary for wine df
summary(wine)
```

The White Wine dataset consists of 13 variables, with almost 5,000 observations.


```{r echo=FALSE, Univariate_Plots1}
# Plot X histogram
ggplot(aes(x = X), data = wine) +
  geom_histogram(bins = length(wine$X)) 

# Plot X line graph with augmented binsize
ggplot(aes(x = X, y = X), data = wine) +
  geom_line()

# Get X statistical values
summary(wine$X)
```


Within the 'wine' df, the 'X' feature acts as an unique ID for each wine given within the df.


```{r echo=FALSE, Univariate_Plots2}
# Plot fixed.acidity histogram
ggplot(aes(x = fixed.acidity), data = wine) +
  geom_histogram(bins = 30)

# Plot fixed.acidity histogram with augmented histogram
ggplot(aes(x = fixed.acidity), data = wine) +
  geom_histogram(binwidth = 0.1) +
  scale_x_continuous(limit = c(4, 11), 
                     breaks = c(seq(4, 11, by = 0.5)))

# Get fixed.acidity statistical values
summary(wine$fixed.acidity)
```


From the dataset documentation:

* fixed acidity (tartaric acid - g / dm^3): most acids involved with wine or fixed or nonvolatile (do not evaporate readily)

The 'fixed.acidity' feature seems to have a wide range of values, normally distrubuted around 6.80. There are some outliers within the wine dataset, which have a 'fixed.acidity' value greater than 11. The 'fixed.acidity' ranges in values from 3.80 to 14.20.


```{r echo=FALSE, Univariate_Plots3}
# Plot volatile.acidity histogram
ggplot(aes(x = volatile.acidity), data = wine) +
  geom_histogram(bins = 30)

# Plot volatile.acidity histogram with augmented histogram
ggplot(aes(x = volatile.acidity), data = wine) +
  geom_histogram(binwidth = 0.01) +
  scale_x_continuous(limit = c(0.1, 0.7), 
                     breaks = c(seq(0.1, 0.7, by = 0.1)))

# Get volatile.acidity statistical values
summary(wine$volatile.acidity)
```


From the dataset documentation:

* volatile acidity (acetic acid - g / dm^3): the amount of acetic acid in wine, which at too high of levels can lead to an unpleasant, vinegar taste

The 'volatile.acidity' feature seems to have a long tail, being positively skewed. The second histogram cuts out the outliers on the few high end values for a more detailed inspection of the bulk of the data. The values for 'volatile.acidity' are centered around the median of 0.26, with a mean of 0.28. The values of 'volatile.acidity' range from 0.08 to 1.10. Based on the description given by the dataset documentation, keeping the 'volatile.acidity' values low would be in the best interest of the consumer.


```{r echo=FALSE, Univariate_Plots4}
# Plot citric.acid histogram
ggplot(aes(x = citric.acid), data = wine) +
  geom_histogram(bins = 30)

# Plot citric.acid histogram with augmented histogram
ggplot(aes(x = citric.acid), data = wine) +
  geom_histogram(binwidth = 0.01) +
  scale_x_continuous(limit = c(0.1, 0.8), 
                     breaks = c(seq(0.1, 0.8, by = 0.1)))

# Get citric.acid statistical values
summary(wine$citric.acid)
```


From the dataset documentation:

* citric acid (g / dm^3): found in small quantities, citric acid can add 'freshness' and flavor to wines

The 'citric.acid' feature seems to have a long tail, bring positively skewed. The values for 'citric.acid' are centered around the median of 0.32, with a mean of 0.33. The values of 'citric.acid' range from 0.00 to 1.66. There is a huge spike in the count of 'citric.acid' at 0.5, this could possibly an industry agreed "sweet spot". Based on the description given by the dataset documentation, keeping the 'citric acid' values high would be in the best interest of the consumer.


```{r echo=FALSE, Univariate_Plots5}
# Plot residual.sugar histogram
ggplot(aes(x = residual.sugar), data = wine) +
  geom_histogram(bins = 30)

# Plot residual.sugar histogram with limits
ggplot(aes(x = residual.sugar), data = wine) +
  geom_histogram(binwidth = 0.1) +
  scale_x_continuous(limits = c(0, 21),
                     breaks = c(seq(0, 21, by = 1)))

# Get residual.sugar statistical values
summary(wine$residual.sugar)
```

From the dataset documentation:

* residual sugar (g / dm^3): the amount of sugar remaining after fermentation stops, it's rare to find wines with less than 1 gram/liter and wines with greater than 45 grams/liter are considered sweet

The 'residual.sugar' feature seems to have a long tail, being positively skewed. The values for residual sugar have high consentrations and spike around the values of 1 and 2, later dropping off. The median for residual sugar is 5.20, with a mean of 6.39.


```{r echo=FALSE, Univariate_Plots6}
# Plot chlorides histogram
ggplot(aes(x = chlorides), data = wine) +
  geom_histogram(bins = 30)

# Plot chlorides histogram with limits
ggplot(aes(x = chlorides), data = wine) +
  geom_histogram(binwidth = 0.001) +
  scale_x_continuous(limits = c(0, 0.2),
                     breaks = c(seq(0, 0.2, by = 0.01)))

# Get residual.sugar statistical values
summary(wine$chlorides)
```


From the dataset documentation:

* chlorides (sodium chloride - g / dm^3): the amount of salt in the wine

The 'chlorides' feature looks normally distributed around the median of 0.043, with a mean of 0.046 but, the data does have a long tail to the right that makes up a small portion of the overall wines. It would make sense for wines to have the smallest amount of salt possible due to wines not being known for their high salinity.


```{r echo=FALSE, Univariate_Plots7}
# Plot free.sulfur.dioxide histogram
ggplot(aes(x = free.sulfur.dioxide), data = wine) +
  geom_histogram(bins = 30)

# Plot free.sulfur.dioxide histogram with limits
ggplot(aes(x = free.sulfur.dioxide), data = wine) +
  geom_histogram(binwidth = 1) +
  scale_x_continuous(limits = c(0, 100),
                     breaks = c(seq(0, 100, by = 5)))

# Get free.sulfur.dioxide statistical values
summary(wine$free.sulfur.dioxide)
```


From the dataset documentation:

* free sulfur dioxide (mg / dm^3): the free form of SO2 exists in equilibrium between molecular SO2 (as a dissolved gas) and bisulfite ion; it prevents microbial growth and the oxidation of wine

The 'free.sulfur.dioxide' feature looks normally distrubuted around the median 34.00, with a mean of 35.31. The histrogram has a long tail to the right, showing outliers with extreme free sulfur dioxide levels. Based on the description of from the dataset documentation, this feature acts as a preservative for the wine.


```{r echo=FALSE, Univariate_Plots8}
# Plot total.sulfur.dioxide histogram
ggplot(aes(x = total.sulfur.dioxide), data = wine) +
  geom_histogram(bins = 30)

# Plot total.sulfur.dioxide histogram with limits
ggplot(aes(x = total.sulfur.dioxide), data = wine) +
  geom_histogram(binwidth = 1) +
  scale_x_continuous(limits = c(0, 300),
                     breaks = c(seq(0, 300, by = 20)))

# Get total.sulfur.dioxide statistical values
summary(wine$total.sulfur.dioxide)
```


From the dataset documentation:

* total sulfur dioxide (mg / dm^3): amount of free and bound forms of S02; in low concentrations, SO2 is mostly undetectable in wine, but at free SO2 concentrations over 50 ppm, SO2 becomes evident in the nose and taste of wine

The 'total.sulfur.dioxide' feature looks normally distrubuted around the median 134.00, with a mean of 138.40. The histrogram has a small tail to the right, showing outliers with extreme total sulfur dioxide levels. Based on the description of from the dataset documentation, the taste of the wine can be affected if the conentration of total sulfur dioxide becomes too high.


```{r echo=FALSE, Univariate_Plots9}
# Plot density histogram
ggplot(aes(x = density), data = wine) +
  geom_histogram(bins = 30)

# Plot density histogram with limits
ggplot(aes(x = density), data = wine) +
  geom_histogram(binwidth = 0.0001) +
  scale_x_continuous(limits = c(0.985, 1.005),
                     breaks = c(seq(0.985, 1.005, by = 0.002)))

# Get density statistical values
summary(wine$density)
```


From the dataset documentation:

* density (g / cm^3): the density of water is close to that of water depending on the percent alcohol and sugar content

The 'density' feature looks normally distrubuted around the median 0.9937, with a mean of 0.9940. The values for wine density within the dataset has some high value outliers, that were cut off when inspecting the second 'density' histogram. Based on the description from the dataset documenation, the wine density is a variable dependent upon the values of alcohol and sugar content.



From the dataset documentation:

* pH: describes how acidic or basic a wine is on a scale from 0 (very acidic) to 14 (very basic); most wines are between 3-4 on the pH scale


From the dataset documentation:

* sulphates (potassium sulphate - g / dm3): a wine additive which can contribute to sulfur dioxide gas (S02) levels, wich acts as an antimicrobial and antioxidant


From the dataset documentation:

* alcohol (% by volume): the percent alcohol content of the wine


From the dataset documentation:

* quality (score between 0 and 10)



### What chemicial properties influence the quality of white wines?

### What chemicial properties influence the alcholol level of white wines?

# Univariate Analysis

### What is the structure of your dataset?
There are 4898 wines of 13 features (x, fixed.acidity, volatile.acidity, citric.acid, residual.sugar, cholorides, free.sulfur.dioxide, total.sulfur.dioxide, density, pH, sulphates, alcohol, quality). All of these variables are given as quanitavive variables.

Other observations:
* The 'quality' distrubution is approximately normal
* The median 'quality' value is 6.0

### What is/are the main feature(s) of interest in your dataset?

### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?

### Did you create any new variables from existing variables in the dataset?

### Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?



# Bivariate Plots Section
```{r echo=FALSE, Bivariate_Plots}

```

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in the dataset?

### Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?

### What was the strongest relationship you found?




# Multivariate Plots Section

```{r echo=FALSE, Multivariate_Plots}

```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. Were there features that strengthened each other in terms of looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the strengths and limitations of your model.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection
