White Wine Dataset - Explore and Summarize Data by James Tooles
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using
# in your analysis in this code chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk.
# This prevents the code from displaying in the knitted HTML output.
# You should set echo=FALSE for all code chunks in your file.

#install.packages('ggplot2', dependencies = T)
#install.packages('knitr', dependencies = T)
#install.packages('dplyr', dependencies = T)
#install.packages('GGally', dependencies = T)

library(ggplot2)
library(knitr)
library(dplyr)
library(GGally)
```


```{r echo=FALSE, Load_the_Data}
# Load the Data
wine <- read.csv(file = 'wineQualityWhites.csv', head = TRUE, sep = ',')
```


# Univariate Plots Section


```{r echo=FALSE, Univariate_Plots0}
# Print info for wine df
str(wine)
 
# Print out head for wine df
head(wine)
  
# Print out summary for wine df
summary(wine)
```

The White Wine dataset consists of 13 variables, with almost 5,000 observations.


```{r echo=FALSE, Univariate_Plots1}
# Plot X histogram
ggplot(aes(x = X), data = wine) +
  geom_histogram(bins = length(wine$X)) 

# Plot X line graph with augmented binsize
ggplot(aes(x = X, y = X), data = wine) +
  geom_line()

# Get X statistical values
summary(wine$X)
```


Within the 'wine' df, the 'X' feature acts as an unique ID for each wine given within the df.


```{r echo=FALSE, Univariate_Plots2}
# Plot fixed.acidity histogram
ggplot(aes(x = fixed.acidity), data = wine) +
  geom_histogram(bins = 30)

# Plot fixed.acidity histogram with augmented histogram
ggplot(aes(x = fixed.acidity), data = wine) +
  geom_histogram(binwidth = 0.1) +
  scale_x_continuous(limit = c(4, 11), 
                     breaks = c(seq(4, 11, by = 0.5)))

# Get fixed.acidity statistical values
summary(wine$fixed.acidity)
```


From the dataset documentation:

* fixed acidity (tartaric acid - g / dm^3): most acids involved with wine or fixed or nonvolatile (do not evaporate readily)

The 'fixed.acidity' feature seems to have a wide range of values, normally distributed around 6.80. There are some outliers within the wine dataset, which have a 'fixed.acidity' value greater than 11. The 'fixed.acidity' ranges in values from 3.80 to 14.20.


```{r echo=FALSE, Univariate_Plots3}
# Plot volatile.acidity histogram
ggplot(aes(x = volatile.acidity), data = wine) +
  geom_histogram(bins = 30)

# Plot volatile.acidity histogram with augmented histogram
ggplot(aes(x = volatile.acidity), data = wine) +
  geom_histogram(binwidth = 0.01) +
  scale_x_continuous(limit = c(0.1, 0.7), 
                     breaks = c(seq(0.1, 0.7, by = 0.1)))

# Get volatile.acidity statistical values
summary(wine$volatile.acidity)
```


From the dataset documentation:

* volatile acidity (acetic acid - g / dm^3): the amount of acetic acid in wine, which at too high of levels can lead to an unpleasant, vinegar taste

The 'volatile.acidity' feature seems to have a long tail, being positively skewed. The second histogram cuts out the outliers on the few high end values for a more detailed inspection of the bulk of the data. The values for 'volatile.acidity' are centered around the median of 0.26, with a mean of 0.28. The values of 'volatile.acidity' range from 0.08 to 1.10. Based on the description given by the dataset documentation, keeping the 'volatile.acidity' values low would be in the best interest of the consumer.


```{r echo=FALSE, Univariate_Plots4}
# Plot citric.acid histogram
ggplot(aes(x = citric.acid), data = wine) +
  geom_histogram(bins = 30)

# Plot citric.acid histogram with augmented histogram
ggplot(aes(x = citric.acid), data = wine) +
  geom_histogram(binwidth = 0.01) +
  scale_x_continuous(limit = c(0.1, 0.8), 
                     breaks = c(seq(0.1, 0.8, by = 0.1)))

# Get citric.acid statistical values
summary(wine$citric.acid)
```


From the dataset documentation:

* citric acid (g / dm^3): found in small quantities, citric acid can add 'freshness' and flavor to wines

The 'citric.acid' feature seems to have a long tail, bring positively skewed. The values for 'citric.acid' are centered around the median of 0.32, with a mean of 0.33. The values of 'citric.acid' range from 0.00 to 1.66. There is a huge spike in the count of 'citric.acid' at 0.5, this could possibly an industry agreed "sweet spot". Based on the description given by the dataset documentation, keeping the 'citric acid' values high would be in the best interest of the consumer.


```{r echo=FALSE, Univariate_Plots5}
# Plot residual.sugar histogram
ggplot(aes(x = residual.sugar), data = wine) +
  geom_histogram(bins = 30)

# Plot residual.sugar histogram with limits
ggplot(aes(x = residual.sugar), data = wine) +
  geom_histogram(binwidth = 0.1) +
  scale_x_continuous(limits = c(0, 21),
                     breaks = c(seq(0, 21, by = 1)))

# Get residual.sugar statistical values
summary(wine$residual.sugar)
```

From the dataset documentation:

* residual sugar (g / dm^3): the amount of sugar remaining after fermentation stops, it's rare to find wines with less than 1 gram/liter and wines with greater than 45 grams/liter are considered sweet

The 'residual.sugar' feature seems to have a long tail, being positively skewed. The values for residual sugar have high concentrations and spike around the values of 1 and 2, later dropping off. The median for residual sugar is 5.20, with a mean of 6.39.


```{r echo=FALSE, Univariate_Plots6}
# Plot chlorides histogram
ggplot(aes(x = chlorides), data = wine) +
  geom_histogram(bins = 30)

# Plot chlorides histogram with limits
ggplot(aes(x = chlorides), data = wine) +
  geom_histogram(binwidth = 0.001) +
  scale_x_continuous(limits = c(0, 0.2),
                     breaks = c(seq(0, 0.2, by = 0.01)))

# Get residual.sugar statistical values
summary(wine$chlorides)
```


From the dataset documentation:

* chlorides (sodium chloride - g / dm^3): the amount of salt in the wine

The 'chlorides' feature looks normally distributed around the median of 0.043, with a mean of 0.046 but, the data does have a long tail to the right that makes up a small portion of the overall wines. It would make sense for wines to have the smallest amount of salt possible due to wines not being known for their high salinity.


```{r echo=FALSE, Univariate_Plots7}
# Plot free.sulfur.dioxide histogram
ggplot(aes(x = free.sulfur.dioxide), data = wine) +
  geom_histogram(bins = 30)

# Plot free.sulfur.dioxide histogram with limits
ggplot(aes(x = free.sulfur.dioxide), data = wine) +
  geom_histogram(binwidth = 1) +
  scale_x_continuous(limits = c(0, 100),
                     breaks = c(seq(0, 100, by = 5)))

# Get free.sulfur.dioxide statistical values
summary(wine$free.sulfur.dioxide)
```


From the dataset documentation:

* free sulfur dioxide (mg / dm^3): the free form of SO2 exists in equilibrium between molecular SO2 (as a dissolved gas) and bisulfite ion; it prevents microbial growth and the oxidation of wine

The 'free.sulfur.dioxide' feature looks normally distributed around the median 34.00, with a mean of 35.31. The histogram has a long tail to the right, showing outliers with extreme free sulfur dioxide levels. Based on the description of from the dataset documentation, this feature acts as a preservative for the wine.


```{r echo=FALSE, Univariate_Plots8}
# Plot total.sulfur.dioxide histogram
ggplot(aes(x = total.sulfur.dioxide), data = wine) +
  geom_histogram(bins = 30)

# Plot total.sulfur.dioxide histogram with limits
ggplot(aes(x = total.sulfur.dioxide), data = wine) +
  geom_histogram(binwidth = 1) +
  scale_x_continuous(limits = c(0, 300),
                     breaks = c(seq(0, 300, by = 20)))

# Get total.sulfur.dioxide statistical values
summary(wine$total.sulfur.dioxide)
```


From the dataset documentation:

* total sulfur dioxide (mg / dm^3): amount of free and bound forms of S02; in low concentrations, SO2 is mostly undetectable in wine, but at free SO2 concentrations over 50 ppm, SO2 becomes evident in the nose and taste of wine

The 'total.sulfur.dioxide' feature looks normally distributed around the median 134.00, with a mean of 138.40. The histogram has a small tail to the right, showing outliers with extreme total sulfur dioxide levels. Based on the description of from the dataset documentation, the taste of the wine can be affected if the concentration of total sulfur dioxide becomes too high.


```{r echo=FALSE, Univariate_Plots9}
# Plot density histogram
ggplot(aes(x = density), data = wine) +
  geom_histogram(bins = 30)

# Plot density histogram with limits
ggplot(aes(x = density), data = wine) +
  geom_histogram(binwidth = 0.0001) +
  scale_x_continuous(limits = c(0.985, 1.005),
                     breaks = c(seq(0.985, 1.005, by = 0.002)))

# Get density statistical values
summary(wine$density)
```


From the dataset documentation:

* density (g / cm^3): the density of water is close to that of water depending on the percent alcohol and sugar content

The 'density' feature looks normally distributed around the median 0.9937, with a mean of 0.9940. The values for wine density within the dataset has some high value outliers, that were cut off when inspecting the second 'density' histogram. Based on the description from the dataset documentation, the wine density is a variable dependent upon the values of alcohol and sugar content.


```{r echo=FALSE, Univariate_Plots10}
# Plot pH histogram
ggplot(aes(x = pH), data = wine) +
  geom_histogram(bins = 30)

# Plot pH histogram with smaller binwidth
ggplot(aes(x = pH), data = wine) +
  geom_histogram(binwidth = 0.01) +
  scale_x_continuous(breaks = c(seq(2.7, 3.8, by = 0.1)))

# Get pH statistical values
summary(wine$pH)
```

From the dataset documentation:

* pH: describes how acidic or basic a wine is on a scale from 0 (very acidic) to 14 (very basic); most wines are between 3-4 on the pH scale

The 'pH' level is known to correlate with the taste, from bitter (high pH = basic) to sour (low pH = acidic). The first 'pH' histogram shows a nice Gaussian distribution of the values for 'pH', centered around the median 3.18, with a mean of 3.188. This is the same shape as the 'quality' feature histogram. The more defined, second histogram, shows some interesting spikes in the counts for different 'pH' levels throughout the histogram.


```{r echo=FALSE, Univariate_Plots11}
# Plot sulphates histogram
ggplot(aes(x = sulphates), data = wine) +
  geom_histogram(bins = 30)

# Plot sulphates histogram with smaller binwidth
ggplot(aes(x = sulphates), data = wine) +
  geom_histogram(binwidth = 0.01) +
  scale_x_continuous(limits = c(0.2, 1.0),
                     breaks = c(seq(0.2, 1.0, by = 0.05)))

# Get sulphates statistical values
summary(wine$sulphates)
```


From the dataset documentation:

* sulphates (potassium sulphate - g / dm3): a wine additive which can contribute to sulfur dioxide gas (S02) levels, which acts as an antimicrobial and antioxidant

The 'sulphates' feature looks normally distributed around the median 0.4700, with a mean of 0.4898. The histogram has a long tail to the right, showing outliers with extreme free sulfur dioxide levels. Based on the description of from the dataset documentation, this feature acts as a preservative for the wine.


```{r echo=FALSE, Univariate_Plots12}
# Plot alcohol histogram
ggplot(aes(x = alcohol), data = wine) +
  geom_histogram(bins = 30)

# Plot alcohol histogram with cut off lower outliers
ggplot(aes(x = alcohol), data = wine) +
  geom_histogram(binwidth = 0.1) +
  scale_x_continuous(limit = c(8.0, 14.2),
                     breaks = c(seq(8.0, 14.2, by = 0.5)))

# Get alcohol statistical values
summary(wine$alcohol)
```


From the dataset documentation:

* alcohol (% by volume): the percent alcohol content of the wine

At the end of the day, wine is an alcoholic product. Perhaps the 'alcohol' content has something to do with the 'quality' value of a wine. The first histogram shows the raw histogram of the 'alcohol' feature. There seem to be a small number of lower outliers.

The second histogram of the 'alcohol' feature removes theses smaller 'alcohol' values from the histogram. Both histograms are positively skewed. There is a noticeable spike at wine alcohol value 9.5. The median of wine alcohol is 10.40, with a mean of 10.51.


```{r echo=FALSE, Univariate_Plots13}
# Plot quality histogram
ggplot(aes(x = quality), data = wine) +
  geom_bar()

# Plot quality histogram with augmentations
ggplot(aes(x = quality), data = wine) +
  geom_bar() +
  scale_x_continuous(limit = c(0, 10),
                     breaks = c(seq(0, 10, by = 1)))

# Get quality statistical values
summary(wine$quality)
```


From the dataset documentation:

* quality (score between 0 and 10)

Starting with a default histogram of the 'quality' feature within the 'wine' dataset resembles a Gaussian distribution of the 'quality' feature, centered at a 'quality' value of 6. By increasing the bin size of the histogram allowed a better visual depiction of the dataset, which is slightly positively skewed. These graphs support the findings within the statistical summary for 'quality' with a median value of 6, and a mean value of 5.878.

The dataset documentation states that the 'quality' feature should range from 0-10, unfortunately all of the wines within this dataset range from 3-9.


# Univariate Analysis

### What is the structure of your dataset?
There are 4898 wines of 13 features (x, fixed.acidity, volatile.acidity, citric.acid, residual.sugar, cholorides, free.sulfur.dioxide, total.sulfur.dioxide, density, pH, sulphates, alcohol, quality). All of these variables are given as quanitavive variables.

* **X**: a unique id identifier for each wine

* **fixed acidity**: most acids involved with wine or fixed or nonvolatile (do not evaporate readily)

* **volatile acidity**: the amount of acetic acid in wine, which at too high of levels can lead to an unpleasant, vinegar taste

* **citric acid**: found in small quantities, citric acid can add 'freshness' and flavor to wines

* **residual sugar**: the amount of sugar remaining after fermentation stops, it's rare to find wines with less than 1 gram/liter and wines with greater than 45 grams/liter are considered sweet

* **chlorides**: the amount of salt in the wine

* **free sulfur dioxide**: the free form of SO2 exists in equilibrium between molecular SO2 (as a dissolved gas) and bisulfite ion; it prevents microbial growth and the oxidation of wine

* **total sulfur dioxide**: amount of free and bound forms of S02; in low concentrations, SO2 is mostly undetectable in wine, but at free SO2 concentrations over 50 ppm, SO2 becomes evident in the nose and taste of wine

* **density**: the density of water is close to that of water depending on the percent alcohol and sugar content

* **pH**: describes how acidic or basic a wine is on a scale from 0 (very acidic) to 14 (very basic); most wines are between 3-4 on the pH scale

* **sulphates**: a wine additive which can contribute to sulfur dioxide gas (S02) levels, wich acts as an antimicrobial and antioxidant

* **alcohol**: the percent alcohol content of the wine

* **quality** (score between 0 and 10)
    * (worst) 0 -------> 10 (best)

Other observations:

* The 'quality' distrubution is approximately normal
* The median 'quality' value is 6.0
* Citric acid has a spike in wine count at a value of 0.5

### What are the main features of interest in your dataset?
The main features of interest are any features that significatly contribute to the quality output value. The goal of this analysis is to determine which features are best for predicting the quality of a wine.

### What other features in the dataset do you think will help support your investigation into your feature of interest?
Volatile acidity, citric acid, residual sugar, chlorides, total sulfur dioxide, pH, and alcohol. All of these features are described in the dataset documentation as to affect the taste, smell, or other measurements of value to consumers such as alcohol content.

The other features such as, fixed acidity, free sulfur dioxide, density, sulphates, do not seem to play a role in determining quality value due to the description given within the dataset documentation. All of thease features are either dependent upon other features of interest, preservatives, or do not show enough variance to be heavily weighted in determining wine quality.

### Did you create any new variables from existing variables in the dataset?
No new variables were created in this tidy wine dataset.

### Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?
The citric acid feature has a spike in wine count at a value of 0.5, this seems odd and interesting and will be investigated for any impact in wine quality within the bivariate section.

All of the features were limited, outliers removed, and/or parsed in more visual detail to see any hidden attributes within the feature.

The quality feature, was shown on a bar graph to highlight the ordinal propery of the feature.


# Bivariate Plots Section


```{r echo=FALSE, Bivariate_Plots0}
# Get correlation matrix values
round(cor(wine), 2)

#Set the seed for reproducible results
set.seed(0)

# Get correlation matrix graph
theme_set(theme_bw(5))
wine_subset <- wine[, c(2:13)]
names(wine_subset)
ggpairs(wine_subset[sample.int(nrow(wine_subset), 500), ])
```


From a subset of the citric acid, residual sugar, and pH do not seem to have strong correlations with quality, but volatile acidity, chlorides, total sulfur dioxide, density, and alcohol are moderately correlated with quality. I want to look closer at scatter plots involving quality and some other variables like volatile acidity, chorides, density, and alcohol.


```{r echo=FALSE, Bivariate_Plots1}
# Plot quality scatter quality vs volatile.acidity
ggplot(aes(x = volatile.acidity, y = quality, colour = quality), data = wine) +
  geom_point(alpha = 0.5, position = position_jitter(width = 0, height = 0.5)) +
  geom_smooth(method = 'lm', color = 'red') +
  scale_x_continuous(breaks = c(seq(0, 1.10, by = 0.1))) +
  scale_y_continuous(breaks = c(seq(0, 10, by = 1)), 
                     limit = c(3, 9)) 

# Plot quality boxplot volatile.acidity vs quality
ggplot(aes(x = quality, y = volatile.acidity, colour = quality), data = wine) +
  geom_boxplot() +
  facet_grid(.~quality) +
  scale_y_continuous(limit = c(0, 0.8),
                     breaks = c(seq(0, 0.8, by = 0.1)))

# Get correlation value
cor.test(wine$quality, wine$volatile.acidity)
```


The high quality wines have a low volatile acidity value, only the sub-par wines have high volatile acidity values. This anaysis only tells us part of the story given the correlation value of -0.195 for quality vs volatile acidity.



```{r echo=FALSE, Bivariate_Plots2}
# Plot quality scatter quality vs chlorides
ggplot(aes(x = chlorides, y = quality, colour = quality), data = wine) +
  geom_point(alpha = 0.5, position = position_jitter(width = 0, height = 0.5)) +
  geom_smooth(method = 'lm', color = 'red') +
  scale_x_continuous(limits = c(0, 0.2),
                     breaks = c(seq(0, 0.2, by = 0.01))) +
  scale_y_continuous(breaks = c(seq(0, 10, by = 1)), 
                     limit = c(3, 9)) 

# Plot quality boxplot chlorides vs quality
ggplot(aes(x = quality, y = chlorides, colour = quality), data = wine) +
  geom_boxplot() +
  facet_grid(.~quality) +
  scale_y_continuous(limits = c(0, 0.15),
                     breaks = c(seq(0, 0.15, by = 0.01)))

# Get correlation value
cor.test(wine$quality, wine$chlorides)
```


The majority of wines have low chlorides values. The high quality wines have a low chlorides value, a large number of average wines have high chlorides values. There are a few outliers on the high quality and low quality wines with a large chloirdes value. This anaysis only tells us part of the story given the correlation value of -0.210 for quality vs chlorides.





```{r echo=FALSE, Bivariate_Plots3}
# Plot quality scatter quality vs total.sulfur.dioxide
ggplot(aes(x = total.sulfur.dioxide, y = quality, colour = quality), data = wine) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = 'lm', color = 'red') +
  scale_x_continuous(limits = c(0, 300),
                     breaks = c(seq(0, 300, by = 20))) +
  scale_y_continuous(breaks = c(seq(0, 10, by = 1)), 
                     limit = c(3, 9)) 

# Plot quality boxplot total.sulfur.dioxide vs quality
ggplot(aes(x = quality, y = total.sulfur.dioxide, colour = quality), data = wine) +
  geom_boxplot() +
  facet_grid(.~quality) +
  scale_y_continuous(limits = c(0, 300),
                     breaks = c(seq(0, 300, by = 20)))

# Get correlation value
cor.test(wine$quality, wine$total.sulfur.dioxide)
```


As the quality of the wines increase, the limits for total sulfur dioxide values seem to converge from 80 to 140, with the low quality wines having a much larger variation of total sulfur dioxide values, about 20 to 280. This anaysis only tells us part of the story given the correlation value of -0.175 for quality vs total sulfur dioxide.


```{r echo=FALSE, Bivariate_Plots4}
# Plot quality scatter quality vs density
ggplot(aes(x = density, y = quality, colour = quality), data = wine) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = 'lm', color = 'red') +
  scale_x_continuous(limits = c(0.985, 1.005),
                     breaks = c(seq(0.985, 1.005, by = 0.002))) +
  scale_y_continuous(breaks = c(seq(0, 10, by = 1)), 
                     limit = c(3, 9)) 

# Plot quality boxplot density vs quality
ggplot(aes(x = quality, y = density, colour = quality), data = wine) +
  geom_boxplot() +
  facet_grid(.~quality) +
  scale_y_continuous(limits = c(0.985, 1.005),
                     breaks = c(seq(0.985, 1.005, by = 0.002)))

# Get correlation value
cor.test(wine$quality, wine$density)
```


The high quality wines have a low density value, only the sub-par wines have high density values. The density and quality of wine has a decent correlation value of -0.307. 



```{r echo=FALSE, Bivariate_Plots5}
# Plot quality scatter quality vs alcohol
ggplot(aes(x = alcohol, y = quality, colour = quality), data = wine) +
  geom_point(alpha = 0.5, position = position_jitter(width = 0, height = 0.5)) +
  geom_smooth(method = 'lm', color = 'red') +
  scale_x_continuous(limit = c(8.0, 14.2),
                     breaks = c(seq(8.0, 14.2, by = 0.5))) +
  scale_y_continuous(breaks = c(seq(0, 10, by = 1)), 
                     limit = c(3, 9)) 

# Plot quality boxplot alcohol vs quality
ggplot(aes(x = quality, y = alcohol, colour = quality), data = wine) +
  geom_boxplot() +
  facet_grid(.~quality) +
  scale_y_continuous(limit = c(8.0, 14.2),
                     breaks = c(seq(8.0, 14.2, by = 0.5)))

# Get correlation value
cor.test(wine$quality, wine$alcohol)
```


The high quality wines have a higher alcohol value, only the sub-par wines have lower alcohol values. Only a small portion of wines have high quality and low alcohol content. The quality and acohol features have the highest correlation yet, at a value of 0.436.


# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in the dataset?

### Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?

### What was the strongest relationship you found?




# Multivariate Plots Section

```{r echo=FALSE, Multivariate_Plots}

```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. Were there features that strengthened each other in terms of looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the strengths and limitations of your model.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection
